---
title: "ggplot2 and patchwork - multiple panes"
description: "Learn how to plot multiple plots on a page."
format:
  html:
    toc: true
    toc-depth: 2
---

# Objective

The goal of this slide set is to show you how to make publication ready graphs and expand on what we have done in the basic plotting series. The goal of this is to use GGPlot and later on I will show you how to use patchwork which allows more flexibility in plot layout.    


# Data for the Exercise

We use a sample M&M dataset. For more sample data files, check out the [Data Files page](99_example_data_files.qmd).

-   [M&M CSV file](data/mms.csv)
-   [M&M Excel file](data/mms.xlsx)

# Load Libraries

Make sure you have installed these packages; if not, run `install.packages("packageName")` separately.

```{r, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Load Libraries ----
# this is done each time you run a script
library(scales)
library(readxl) # read in excel files
library(janitor) # clean up excel imports
library(patchwork) # multipanel graphs
library(tidyverse) # dplyr and piping and ggplot etc
```

So now we have seen how to look at the data    
What if we wanted to modify the data in terms of columns or rows    
     
Making graphs this way can get a bit cumbersome as you might imagine.     
This is because the data is in what we call wide format     
The long format is the format often used for Anovas and other stats      
We will go over how to do this later but for now lets just look at the file     

# Read in the Data

We read the lakes data for this one.

```{r, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
# This code runs but isn't shown in the rendered document.
lakes.df <- read_csv("data/reduced_lake_long_genus_species.csv")
head(lakes.df)
```
# Patchwork graphing-----
# https://github.com/thomasp85/patchwork
# we can have a lot more control over our plots if we want

```{r, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
# plot of Willis Lake ------
willis.plot <- lakes.df %>%
  filter(lake_name=="Willis" & str_detect(genus_species, "Daphnia")) %>% 
  ggplot(aes(date, org_l, color=genus_species)) + # sometimes necessary is , group = group
  geom_point()+
  geom_line() +
  labs(x = "Date", y = expression(bold("Animals (No. L"^-1*")"))) +
  scale_x_date(date_breaks = "6 month",
               limits = as_date(c('1994-06-01', '2006-12-31')),
               labels=date_format("%Y-%m-%d"), expand=c(0,0))  +
  theme(axis.text.x = element_text(size=12, face="bold", angle=45, hjust=1)) 
willis.plot
```

## Plot Grass Lake


```{r, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
grass.plot <- lakes.df %>%
  filter(lake_name=="Grass" & str_detect(genus_species, "Daphnia")) %>% 
  ggplot(aes(date, org_l, color=genus_species)) + # sometimes necessary is , group = group
  geom_point()+
  geom_line() +
  labs(x = "Date", y = expression(bold("Animals (No. L"^-1*")"))) +
  scale_x_date(date_breaks = "6 month",
               limits = as_date(c('1994-06-01', '2006-12-31')),
               labels=date_format("%Y-%m-%d"), expand=c(0,0))  +
  theme(axis.text.x = element_text(size=12, face="bold", angle=45, hjust=1)) 
grass.plot
```


```{r, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
# plot of Indian Lake ------
indian.plot <- lakes.df %>%
  filter(lake_name=="Indian" & str_detect(genus_species, "Daphnia")) %>% 
  ggplot(aes(date, org_l, color=genus_species)) + # sometimes necessary is , group = group
  geom_point()+
  geom_line() +
  labs(x = "Date", y = expression(bold("Animals (No. L"^-1*")"))) +
  scale_x_date(date_breaks = "6 month",
               limits = as_date(c('1994-06-01', '2006-12-31')),
               labels=date_format("%Y-%m-%d"), expand=c(0,0))  +
  theme(axis.text.x = element_text(size=12, face="bold", angle=45, hjust=1)) 
indian.plot

```


```{r, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
# Now we can use patchwork to combine files
# Lets look at the plots in one format
grass.plot +
willis.plot +
indian.plot +
plot_layout(ncol = 1)

```

```{r, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# here is another format
grass.plot +
willis.plot +
indian.plot +
plot_layout(ncol = 1,
              heights=c(4,2,1))
```


```{r, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
# you might want to turn off legends and axes
# you can do this in the theme statements
grass.plot + theme(axis.title.x=element_blank(), axis.text.x=element_blank(),
                   legend.position = "none") + 
willis.plot + theme(axis.title.x=element_blank(), axis.text.x=element_blank(),
                    legend.position = "none") +
indian.plot + theme(axis.title.x=element_blank(), axis.text.x=element_blank(),
                    legend.position = "bottom") +
plot_layout(ncol = 1,
            heights=c(4,2,1))

```



```{r, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
# the plots can get as fancy as you would like

grass.plot +  theme(legend.position = "bottom") + {
  willis.plot + theme(axis.title.x=element_blank(), axis.text.x=element_blank(),
                      legend.position = "none") +
  indian.plot +  theme(legend.position = "none") +
    plot_layout(ncol=1) } +
  plot_layout(ncol = 2)

```



```{r, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# You can also do this without brackets
(grass.plot  + theme(axis.title.x=element_blank(), axis.text.x=element_blank(),
                     legend.position = "none") |
willis.plot + theme(axis.title.x=element_blank(), axis.text.x=element_blank(),
                      legend.position = "none")) /
indian.plot +   theme(legend.position = "bottom") 
```

