---
title: "T-Test"
description: " "
format:
  html:
    toc: true
    toc-depth: 2
---

# Objective

The goal of this page is to learn to perform a t-test

## T Tests

##Load libraries
We will read in the main files and load the libraries as we have worked with so far.  
```{r, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
# One new package for summary stats
#install.packages("broom")
# install.packages("GGally")
# install.packages("car")
# install.packages("gvlma")


# load the libraries each time you restart R
library(tidyverse)
library(lubridate)
library(scales)
library(skimr)
library(janitor)
library(patchwork)
# library(reshape2)
library(broom)
library(GGally)
library(corrplot)
library(car)
```

## Read in file
```{r, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
# read in the file
iris.df <- read_csv("data/iris.csv") %>%
  clean_names() %>%
  remove_empty(c("rows", "cols")) 

glimpse(iris.df)
```


###Summary Statistics for the better look
###So this is a lot different than thinking about data from excel    
Lets try to do the summary stats on the data now and see how it differs
```{r, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
# the data you want to look at
skim(iris.df)
```
   
```{r, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
iris.df %>% group_by(species) %>% skim_to_wide()
```

## Wide to long format
```{r, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
# this will add an index to the dataframe so you know what individual is which
iris_long.df <- iris.df %>% 
  mutate(individual = row_number()) %>% 
  gather(trait, measure, -species, - individual)

head(iris_long.df)
```


## Outliers
```{r, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
# Box Plots of data
iris_long.df %>% group_by(species, trait) %>% 
  ggplot( aes(x = trait, y = measure, color = species, fill=species))+
  geom_boxplot(aes(alpha=0.3)) 
```

These look good for the most part with only one or two significant. Your choice to transform or not.       

##Factors     
Make sure the categorical variable is a factor    
Rearrange the order of groups so that control group is first, followed by treatment groups.     
you can do this by:
```{r, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
# Make Factors from the different levels long way
# iris_long.df$species <- as.factor(iris_long.df$species)  
# iris_long.df$trait <- as.factor(iris_long.df$trait) 

# Make Factors dplyr
iris_long.df <- iris_long.df %>%
  mutate(
    sex = factor(species, 
    labels = c("setosa", "versicolor", "virginica")),
    trait = factor(trait, 
    labels = c("sepal_length", "sepal_width", "petal_length", "petal_width"))
  )
```

## Test for normality of data and using the broom package     
So I think this is premature but some people like to test the normality of the data but really you should be assessing the normality of the residuals. But here it goes...

```{r, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
# turn off scientific notaton
options(scipen = 999)
# to turn back on 
#options(scipen = 0)

# Test for normality of each group and store in shapirowilktests
# This uses the broom package to get clean output of the test 
iris_long.df %>% group_by(species, trait) %>% do(tidy(shapiro.test(.$measure)))

#You can do this on all variables faster with if there was only one grouping
# tapply(iris_long.df$measure, iris_long.df$species, shapiro.test)
```  

##Test for homogenetiy of variances       

Best to use the Levenes test compared to the Bartlet test    
uses the car package      

http://www.cookbook-r.com/Statistical_analysis/Homogeneity_of_variance/   

Note however that this is doing the homogeneity test on all traits and not each one which is what you really should do and is not often done. Need to check on this though.

```{r, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
#Test for homogeneity of variances by groups
leveneTest(sepal_length ~ species, data=iris.df)

#Compared to the Bartlet test
bartlett.test(sepal_length ~ species, data=iris.df)

```

Create a subset dataframe    
```{r, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
sepal_length.df <- iris_long.df %>%
  filter(trait =="sepal_length") %>%
  filter(species != "setosa")
```

If p-value >= 0.05, use var.equal=TRUE below
```{r, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
t.test(measure ~ species, data=sepal_length.df,
       var.equal=TRUE,
       conf.level=0.95)
```

```{r, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
t.test(measure ~ species, data=sepal_length.df,
       var.equal=FALSE,
       conf.level=0.95)
```

Boxplot of data
```{r, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
sepal_length.df %>%
  ggplot(aes(species, measure, fill = species)) +
  geom_boxplot()
```

Mean and standard error plot
```{r, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
sepal_length.df %>% 
ggplot(aes(species, color=species)) + 
  stat_summary(aes(y = measure),
               fun.y = mean, na.rm = TRUE,
               geom = "point",
               size = 3) + 
  stat_summary(aes(y = measure),
               fun.data = mean_se, na.rm = TRUE,
               geom = "errorbar",
               width = 0.2) +
  labs(x = "Species", y = "Length") +
  scale_color_manual(name = "Species", 
                     values = c("blue", "red"),
                     labels = c("Versicolor", "Virginica"))
```


   

   
   
   